# Preface {-}

## Why this book?

Capture-recapture. HMM. And Bayes.

## Who should read this book?

This book is aimed at two main audiences:

-   Beginners, but comfortable using R to do data analysis, and loops. 

-   But also advanced analysts, frequentist framework, for Bayesian. Also others for HMM and open the full flexibility of framework. Uncertainty. 

## What will you learn?

The book is divided into two parts. Give description.

## What won't you learn?

No math ou presque. Don't cover Bayesian statistics, or HMM fully. Suggested reading. There are also a number of important topics specific to capture-recapture that I don't cover: closed-population, and spatial capture-recapture. Refer to books. 

## Prerequisites {#prerequisites}

This book uses primarily the R package NIMBLE, so you need to at least install R and the NIMBLE package. Also a bunch of R packages. You can install them all at once by running:

```{r, echo = FALSE, cache = FALSE}
deps <- desc::desc_get_deps()
pkgs <- sort(deps$package[deps$type == "Imports"])
pkgs2 <- strwrap(paste(encodeString(pkgs, quote = '"'), collapse = ", "), exdent = 2)
install <- paste0(
  "install.packages(c(\n  ", 
  paste(pkgs2, "\n", collapse = ""), 
  "))"
)
```

```{r code = install, eval = FALSE}
```

## Acknowledgements

This book was written in the open and chapters were advertised on twitter when complete. Read drafts, fixed typos, and suggested improvements. I'm deeply grateful for their help.

## How this book was written

This book was written in [RStudio](http://www.rstudio.com/ide/) using [bookdown](http://bookdown.org/). The [website](https://oliviergimenez.github.io/banana-book) is hosted with [GitHub Pages](https://pages.github.com/), and automatically updated after every push by [Github Actions](https://github.com/features/actions). The complete source is available from [GitHub](https://github.com/oliviergimenez/banana-book).

This version of the book was built with `r R.version.string` and the following packages:

```{r, echo = FALSE, results="asis"}
pkgs <- sessioninfo::package_info(pkgs, dependencies = FALSE)
df <- tibble(
  package = pkgs$package,
  version = pkgs$ondiskversion,
  source = gsub("@", "\\\\@", pkgs$source)
)
knitr::kable(df, format = "markdown")
```

```{r, echo = FALSE}
ruler <- function(width = getOption("width")) {
  x <- seq_len(width)
  y <- dplyr::case_when(
    x %% 10 == 0 ~ as.character((x %/% 10) %% 10),
    x %% 5 == 0  ~ "+",
    TRUE         ~ "-"
  )
  cat(y, "\n", sep = "")
  cat(x %% 10, "\n", sep = "")
}
```

```{r, include = FALSE}
ruler()
```
