---
title: "Chapter 3: Hidden Markov models"
date: "last updated: `r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE, echo=FALSE, message=FALSE, warning=FALSE}
# R options
options(width = 60)
# chunk options
knitr::opts_chunk$set(
  comment = "##",
  collapse = TRUE,
  warning = FALSE,
  message = FALSE
  )
```

## Introduction

## Longitudinal data

```{r}
library(nimble)
# 1 = alive, 2 = dead
nind <- 57
nocc <- 5
phi <- 0.8 # survival probability
delta <- c(1,0) # (Pr(alive at t = 1), Pr(dead at t = 1))
Gamma <- matrix(NA, 2, 2) # transition matrix
Gamma[1,1] <- phi      # Pr(alive t -> alive t+1)
Gamma[1,2] <- 1 - phi  # Pr(alive t -> dead t+1)
Gamma[2,1] <- 0        # Pr(dead t -> alive t+1)
Gamma[2,2] <- 1        # Pr(dead t -> dead t+1)
z <- matrix(NA, nrow = nind, ncol = nocc)
set.seed(2022)
for (i in 1:nind){
  z[i,1] <- rcat(n = 1, prob = delta) # 1 for sure
  for (t in 2:nocc){
    z[i,t] <- rcat(n = 1, prob = Gamma[z[i,t-1],1:2]) 
  }
}
colnames(z) <- paste0("winter ", 1:nocc)
head(z, n = 20)
```

## A Markov model for longitudinal data

### Assumptions 

### Transition matrix 

### Initial states 

### Likelihood 

### Example 

## Bayesian formulation 

## NIMBLE implementation 

```{r}
library(nimble)
library(MCMCvis)

markov.survival <- nimbleCode({
  phi ~ dunif(0, 1) # prior
  delta[1] <- 1          # Pr(alive t = 1) = 1
  delta[2] <- 0          # Pr(dead t = 1) = 0
  gamma[1,1] <- phi      # Pr(alive t -> alive t+1)
  gamma[1,2] <- 1 - phi  # Pr(alive t -> dead t+1)
  gamma[2,1] <- 0        # Pr(dead t -> alive t+1)
  gamma[2,2] <- 1        # Pr(dead t -> dead t+1)
  # likelihood
  for (i in 1:N){ # loop over individual i
    z[i,1] ~ dcat(delta[1:2]) # t = 1
    for (j in 2:T){ # loop over time t
      z[i,j] ~ dcat(gamma[z[i,j-1], 1:2]) # t = 2,...,T
    } # t
  } # i
})

my.constants <- list(N = 57, T = 5)
my.data <- list(z = z)

initial.values <- function() list(phi = runif(1,0,1))

parameters.to.save <- c("phi")

n.iter <- 5000
n.burnin <- 1000
n.chains <- 2

mcmc.output <- nimbleMCMC(code = markov.survival,
                          constants = my.constants,
                          data = my.data,
                          inits = initial.values,
                          monitors = parameters.to.save,
                          niter = n.iter,
                          nburnin = n.burnin,
                          nchains = n.chains)

MCMCsummary(mcmc.output, round = 2)
```


## Hidden Markov models 

### Capture-recapture data

```{r}
library(nimble)
set.seed(2022) # for reproducibility
nocc <- 5 # nb of winters or sampling occasions
nind <- 57 # nb of animals
p <- 0.6 # detection prob
phi <- 0.8 # survival prob
# Vector of initial states probabilities
delta <- c(1,0) # all individuals are alive in first winter
# Transition matrix
Gamma <- matrix(NA, 2, 2)
Gamma[1,1] <- phi      # Pr(alive t -> alive t+1)
Gamma[1,2] <- 1 - phi  # Pr(alive t -> dead t+1)
Gamma[2,1] <- 0        # Pr(dead t -> alive t+1)
Gamma[2,2] <- 1        # Pr(dead t -> dead t+1)
# Observation matrix 
Omega <- matrix(NA, 2, 2)
Omega[1,1] <- 1 - p      # Pr(alive t -> non-detected t)
Omega[1,2] <- p          # Pr(alive t -> detected t)
Omega[2,1] <- 1          # Pr(dead t -> non-detected t)
Omega[2,2] <- 0          # Pr(dead t -> detected t)
# Matrix of states
z <- matrix(NA, nrow = nind, ncol = nocc)
y <- z
y[,1] <- 2 # all individuals are detected in first winter
for (i in 1:nind){
  z[i,1] <- rcat(n = 1, prob = delta) # 1 for sure
  for (t in 2:nocc){
    # state at t given state at t-1
    z[i,t] <- rcat(n = 1, prob = Gamma[z[i,t-1],1:2]) 
    # observation at t given state at t
    y[i,t] <- rcat(n = 1, prob = Omega[z[i,t],1:2]) 
  }
}
y
y <- y - 1 # non-detection = 0, detection = 1
head(y)
```

### Observation matrix

### Hidden Markov model

### Likelihood

## Fitting HMM with NIMBLE

```{r}
library(nimble)
library(MCMCvis)

hmm.survival <- nimbleCode({
  phi ~ dunif(0, 1) # prior survival
  p ~ dunif(0, 1) # prior detection
  # likelihood
  delta[1] <- 1          # Pr(alive t = 1) = 1
  delta[2] <- 0          # Pr(dead t = 1) = 0
  gamma[1,1] <- phi      # Pr(alive t -> alive t+1)
  gamma[1,2] <- 1 - phi  # Pr(alive t -> dead t+1)
  gamma[2,1] <- 0        # Pr(dead t -> alive t+1)
  gamma[2,2] <- 1        # Pr(dead t -> dead t+1)
  omega[1,1] <- 1 - p    # Pr(alive t -> non-detected t)
  omega[1,2] <- p        # Pr(alive t -> detected t)
  omega[2,1] <- 1        # Pr(dead t -> non-detected t)
  omega[2,2] <- 0        # Pr(dead t -> detected t)
  for (i in 1:N){
    z[i,1] ~ dcat(delta[1:2])
    for (j in 2:T){
      z[i,j] ~ dcat(gamma[z[i,j-1], 1:2])
      y[i,j] ~ dcat(omega[z[i,j], 1:2])
    }
  }
})

my.constants <- list(N = nrow(y), T = 5)

my.data <- list(y = y + 1)

zinits <- y + 1 # non-detection -> alive
zinits[zinits == 2] <- 1 # dead -> alive
initial.values <- function() list(phi = runif(1,0,1),
                                  p = runif(1,0,1),
                                  z = zinits)

parameters.to.save <- c("phi", "p")

n.iter <- 5000
n.burnin <- 1000
n.chains <- 2

start_time <- Sys.time()
mcmc.output <- nimbleMCMC(code = hmm.survival,
                          constants = my.constants,
                          data = my.data,
                          inits = initial.values,
                          monitors = parameters.to.save,
                          niter = n.iter,
                          nburnin = n.burnin,
                          nchains = n.chains)
end_time <- Sys.time()
end_time - start_time

MCMCsummary(mcmc.output, round = 2)
```

## Marginalization 

### Brute-force approach

### Forward algorithm 

### NIMBLE implementation 

#### Do it yourself

```{r}
library(nimble)
library(MCMCvis)

dHMM <- nimbleFunction(
  run = function(x = double(1), 
                 probInit = double(1),
                 probObs = double(2),
                 probTrans = double(2),
                 len = double(0, default = 0),
                 log = integer(0, default = 0)) {
    alpha <- probInit[1:2]
    for (t in 2:len) {
      alpha[1:2] <- (alpha[1:2] %*% probTrans[1:2,1:2]) * probObs[1:2,x[t]]
    }
    logL <- log(sum(alpha[1:2]))
    returnType(double(0))
    if (log) return(logL)
    return(exp(logL))
  }
)

rHMM <- nimbleFunction(
  run = function(n = integer(),
                 probInit = double(1),
                 probObs = double(2),
                 probTrans = double(2),
                 len = double(0, default = 0)) {
    returnType(double(1))
    z <- numeric(len)
    z[1] <- rcat(n = 1, prob = probInit[1:2]) # all individuals alive at t = 0
    y <- z
    y[1] <- 2 # all individuals are detected at t = 0
    for (t in 2:len){
      # state at t given state at t-1
      z[t] <- rcat(n = 1, prob = probTrans[z[t-1],1:2]) 
      # observation at t given state at t
      y[t] <- rcat(n = 1, prob = probObs[z[t],1:2]) 
    }
    return(y)
  })

assign('dHMM', dHMM, .GlobalEnv)
assign('rHMM', rHMM, .GlobalEnv)

# code
hmm.survival <- nimbleCode({
  phi ~ dunif(0, 1) # prior survival
  p ~ dunif(0, 1) # prior detection
  # likelihood
  delta[1] <- 1          # Pr(alive t = 1) = 1
  delta[2] <- 0          # Pr(dead t = 1) = 0
  gamma[1,1] <- phi      # Pr(alive t -> alive t+1)
  gamma[1,2] <- 1 - phi  # Pr(alive t -> dead t+1)
  gamma[2,1] <- 0        # Pr(dead t -> alive t+1)
  gamma[2,2] <- 1        # Pr(dead t -> dead t+1)
  omega[1,1] <- 1 - p    # Pr(alive t -> non-detected t)
  omega[1,2] <- p        # Pr(alive t -> detected t)
  omega[2,1] <- 1        # Pr(dead t -> non-detected t)
  omega[2,2] <- 0        # Pr(dead t -> detected t)
  for (i in 1:N){
    y[i,1:T] ~ dHMM(probInit = delta[1:2], 
                    probObs = omega[1:2,1:2], # observation matrix
                    probTrans = gamma[1:2,1:2], # transition matrix
                    len = T) # nb of sampling occasions
  }
})
# constants
my.constants <- list(N = nrow(y), T = 5)

# data
my.data <- list(y = y + 1)
# initial values - no need to specify values for z anymore
initial.values <- function() list(phi = runif(1,0,1),
                                  p = runif(1,0,1))


# my.constants <- list(N = 1, T = 5)
# my.data <- list(y = matrix(y[1,], nrow = 1) + 1)
# HMM_model <- nimbleModel(code = hmm.survival,
#                          constants = my.constants,
#                          data = my.data,
#                          inits = list(phi = 0.5, p = 0.5))
# HMM_model$calculate()
# 
# alpha <- c(1,0)
# phi <- p <- 0.5
# probObs <- probTrans <- matrix(numeric(4), nrow = 2)
# probTrans[1,1] <- phi      # Pr(alive t -> alive t+1)
# probTrans[1,2] <- 1 - phi  # Pr(alive t -> dead t+1)
# probTrans[2,1] <- 0        # Pr(dead t -> alive t+1)
# probTrans[2,2] <- 1        # Pr(dead t -> dead t+1)
# probObs[1,1] <- 1 - p    # Pr(alive t -> non-detected t)
# probObs[1,2] <- p        # Pr(alive t -> detected t)
# probObs[2,1] <- 1        # Pr(dead t -> non-detected t)
# probObs[2,2] <- 0        # Pr(dead t -> detected t)
# x <- c(1,0,0,0,0) + 1
# logL <- 0
# #alpha[1:2] <- alpha[1:2] * probObs[1:2,x[1]]
# alpha[1:2] <- alpha[1:2]
# for (t in 2:5) {
#   alpha[1:2] <- (alpha[1:2] %*% probTrans[1:2,1:2]) * probObs[1:2,x[t]]
# }
# logL <- log(sum(alpha[1:2]))
# logL

# parameters to save
parameters.to.save <- c("phi", "p")

# MCMC details
n.iter <- 5000
n.burnin <- 1000
n.chains <- 2

start_time <- Sys.time()
mcmc.output <- nimbleMCMC(code = hmm.survival,
                          constants = my.constants,
                          data = my.data,
                          inits = initial.values,
                          monitors = parameters.to.save,
                          niter = n.iter,
                          nburnin = n.burnin,
                          nchains = n.chains)
end_time <- Sys.time()
end_time - start_time

MCMCsummary(mcmc.output, round = 2)
```

#### Do it with `nimbleEcology`

```{r}
library(nimbleEcology)
library(MCMCvis)

# code
hmm.survival <- nimbleCode({
  phi ~ dunif(0, 1) # prior survival
  p ~ dunif(0, 1) # prior detection
  # likelihood
  delta[1] <- 1          # Pr(alive t = 1) = 1
  delta[2] <- 0          # Pr(dead t = 1) = 0
  gamma[1,1] <- phi      # Pr(alive t -> alive t+1)
  gamma[1,2] <- 1 - phi  # Pr(alive t -> dead t+1)
  gamma[2,1] <- 0        # Pr(dead t -> alive t+1)
  gamma[2,2] <- 1        # Pr(dead t -> dead t+1)
  omega[1,1,1] <- 0        # Pr(alive first -> non-detected first)
  omega[1,2,1] <- 1        # Pr(alive first -> detected first)
  omega[2,1,1] <- 1        # Pr(dead first -> non-detected first)
  omega[2,2,1] <- 0        # Pr(dead first -> detected first)
  for (t in 2:5){
    omega[1,1,t] <- 1 - p    # Pr(alive t -> non-detected t)
    omega[1,2,t] <- p        # Pr(alive t -> detected t)
    omega[2,1,t] <- 1        # Pr(dead t -> non-detected t)
    omega[2,2,t] <- 0        # Pr(dead t -> detected t)
  }
  for (i in 1:N){
    y[i,1:5] ~ dHMMo(init = delta[1:2], # vector of initial state probabilities
                     probObs = omega[1:2,1:2,1:5], # observation matrix
                     probTrans = gamma[1:2,1:2], # transition matrix
                     len = 5, # nb of sampling occasions
                     checkRowSums = 0) # skip validity checks
  }
})

# constants
my.constants <- list(N = nrow(y))

# data
my.data <- list(y = y + 1)

# initial values - no need to specify values for z anymore
initial.values <- function() list(phi = runif(1,0,1),
                                  p = runif(1,0,1))

# my.constants <- list(N = 1)
# my.data <- list(y = matrix(c(1,0,0,0,0), nrow = 1) + 1)
# HMM_model <- nimbleModel(code = hmm.survival,
#                          constants = my.constants,
#                          data = my.data,
#                          inits = list(phi = 0.5, p = 0.5))
# HMM_model$calculate()
# 
# pi <- c(1,0)
# logL <- 0
# nObsClasses <- 2
# phi <- p <- 0.5
# probObs <- array(numeric(4), dim = c(2,2,5))
# probTrans <- matrix(numeric(4), nrow = 2)
# probTrans[1,1] <- phi      # Pr(alive t -> alive t+1)
# probTrans[1,2] <- 1 - phi  # Pr(alive t -> dead t+1)
# probTrans[2,1] <- 0        # Pr(dead t -> alive t+1)
# probTrans[2,2] <- 1        # Pr(dead t -> dead t+1)
# probObs[1,1,1] <- 0    # Pr(alive t -> non-detected t)
# probObs[1,2,1] <- 1        # Pr(alive t -> detected t)
# probObs[2,1,1] <- 1        # Pr(dead t -> non-detected t)
# probObs[2,2,1] <- 0        # Pr(dead t -> detected t)
# for (t in 2:5){
#   probObs[1,1,t] <- 1 - p    # Pr(alive t -> non-detected t)
#   probObs[1,2,t] <- p        # Pr(alive t -> detected t)
#   probObs[2,1,t] <- 1        # Pr(dead t -> non-detected t)
#   probObs[2,2,t] <- 0        # Pr(dead t -> detected t)
# }
# x <- c(1,0,0,0,0) + 1
# logL <- 0
# for (t in 1:5) {
#   Zpi <- probObs[, x[t], t] * pi
#   sumZpi <- sum(Zpi)
#   logL <- logL + log(sumZpi)
#   if (t != 5) 
#     pi <- ((Zpi %*% probTrans)/sumZpi)[1, ]
# }
# logL

# parameters to save
parameters.to.save <- c("phi", "p")

# MCMC details
n.iter <- 5000
n.burnin <- 1000
n.chains <- 2

start_time <- Sys.time()
mcmc.output <- nimbleMCMC(code = hmm.survival,
                          constants = my.constants,
                          data = my.data,
                          inits = initial.values,
                          monitors = parameters.to.save,
                          niter = n.iter,
                          nburnin = n.burnin,
                          nchains = n.chains)
end_time <- Sys.time()
end_time - start_time

MCMCsummary(mcmc.output, round = 2)
```

## Pooled encounter histories

### Do it yourself

```{r}
library(nimble)
library(MCMCvis)
library(tidyverse)

dHMMpooled <- nimbleFunction(
  run = function(x = double(1), 
                 probInit = double(1),
                 probObs = double(2),
                 probTrans = double(2),
                 len = double(0),
                 size = double(0),
                 log = integer(0, default = 0)) {
    alpha <- probInit[1:2]
    for (t in 2:len) {
      alpha[1:2] <- (alpha[1:2] %*% probTrans[1:2,1:2]) * probObs[1:2,x[t]]
    }
    logL <- log(sum(alpha[1:2])) * size
    returnType(double(0))
    if (log) return(logL)
    return(exp(logL))
  }
)

rHMMpooled <- nimbleFunction(
  run = function(n = integer(),
                 probInit = double(1),
                 probObs = double(2),
                 probTrans = double(2),
                 len = double(0),
                 size = double(0)) {
    returnType(double(1))
    z <- numeric(len)
    z[1] <- rcat(n = 1, prob = probInit[1:2]) # all individuals alive at t = 0
    y <- z
    y[1] <- 2 # all individuals are detected at t = 0
    for (t in 2:len){
      # state at t given state at t-1
      z[t] <- rcat(n = 1, prob = probTrans[z[t-1],1:2]) 
      # observation at t given state at t
      y[t] <- rcat(n = 1, prob = probObs[z[t],1:2]) 
    }
    return(y)
  })

assign('dHMMpooled', dHMMpooled, .GlobalEnv)
assign('rHMMpooled', rHMMpooled, .GlobalEnv)

hmm.survival <- nimbleCode({
  phi ~ dunif(0, 1) # prior survival
  p ~ dunif(0, 1) # prior detection
  # likelihood
  delta[1] <- 1          # Pr(alive t = 1) = 1
  delta[2] <- 0          # Pr(dead t = 1) = 0
  gamma[1,1] <- phi      # Pr(alive t -> alive t+1)
  gamma[1,2] <- 1 - phi  # Pr(alive t -> dead t+1)
  gamma[2,1] <- 0        # Pr(dead t -> alive t+1)
  gamma[2,2] <- 1        # Pr(dead t -> dead t+1)
  omega[1,1] <- 1 - p    # Pr(alive t -> non-detected t)
  omega[1,2] <- p        # Pr(alive t -> detected t)
  omega[2,1] <- 1        # Pr(dead t -> non-detected t)
  omega[2,2] <- 0        # Pr(dead t -> detected t)
  for (i in 1:N){
    y[i,1:T] ~ dHMMpooled(probInit = delta[1:2], 
                          probObs = omega[1:2,1:2], # observation matrix
                          probTrans = gamma[1:2,1:2], # transition matrix
                          len = T, # nb of sampling occasions
                          size = size[i]) # number of individuals with encounter history i
  }
})

y_pooled <- y %>% 
  as_tibble() %>% 
  group_by_all() %>% # group
  summarise(size = n()) %>% # count
  relocate(size) %>% # put size in front
  arrange(-size) %>% # sort along size
  as.matrix()
y_pooled

my.constants <- list(N = nrow(y_pooled), T = 5, size = y_pooled[,'size'])
my.data <- list(y = y_pooled[,-1] + 1) # delete size from dataset
initial.values <- function() list(phi = runif(1,0,1),
                                  p = runif(1,0,1))
parameters.to.save <- c("phi", "p")
n.iter <- 5000
n.burnin <- 1000
n.chains <- 2
start_time <- Sys.time()
mcmc.output <- nimbleMCMC(code = hmm.survival,
                          constants = my.constants,
                          data = my.data,
                          inits = initial.values,
                          monitors = parameters.to.save,
                          niter = n.iter,
                          nburnin = n.burnin,
                          nchains = n.chains)
end_time <- Sys.time()
end_time - start_time

MCMCsummary(mcmc.output, round = 2)
```

### Do it with `nimbleEcology`

```{r}
library(nimble)
library(MCMCvis)
library(tidyverse)

# get functions in nimbleEcology GitHub <https://github.com/nimble-dev/nimbleEcology/blob/master/R/dHMM.R>
# and add size

dHMMopooled <- nimbleFunction(
  run = function(x = double(1),    ## Observed capture (state) history
                 init = double(1),##
                 probObs = double(3),
                 probTrans = double(2),
                 len = double(0, default = 0),## length of x (needed as a separate param for rDHMM)
                 size = double(0),
                 checkRowSums = double(0, default = 1),
                 log = integer(0, default = 0)) {
    if (length(x) != len) stop("In dHMMo: Argument len must be length of x or 0.")
    if (dim(probObs)[1] != dim(probTrans)[1]) stop("In dHMMo: In dHMM: Length of dimension 1 in probObs must equal length of dimension 1 in probTrans.")
    if (dim(probTrans)[1] != dim(probTrans)[2]) stop("In dHMMo: probTrans must be a square matrix.")
    if (dim(probObs)[3] != len) {
      if (dim(probObs)[3] == 1) stop("In dHMMo: Time dimension of probObs must match length of data. Did you mean dHMM?")
      stop("In dHMMo: Length of time dimension of probObs must match length of data.")
    }
    if (abs(sum(init) - 1) > 1e-6) stop("In dHMMo: Initial probabilities must sum to 1.")

    if (checkRowSums) {
      transCheckPasses <- TRUE
      for (i in 1:dim(probTrans)[1]) {
        thisCheckSum <- sum(probTrans[i,])
        if (abs(thisCheckSum - 1) > 1e-6) {
          ## Compilation doesn't support more than a simple string for stop()
          ## so we provide more detail using a print().
          print("In dHMMo: Problem with sum(probTrans[i,]) with i = ", i, ". The sum should be 1 but is ", thisCheckSum)
          transCheckPasses <- FALSE
        }
      }
      obsCheckPasses <- TRUE
      for (i in 1:dim(probObs)[1]) {
        for (k in 1:dim(probObs)[3]) {
          thisCheckSum <- sum(probObs[i,,k])
          if (abs(thisCheckSum - 1) > 1e-6) {
            print("In dHMMo: Problem with sum(probObs[i,,k]) with i = ", i, " k = " , k, ". The sum should be 1 but is ", thisCheckSum)
            obsCheckPasses <- FALSE
          }
        }
      }
      if(!(transCheckPasses | obsCheckPasses))
        stop("In dHMMo: probTrans and probObs were not specified correctly.  Probabilities in each row (second dimension) must sum to 1.")
      if(!transCheckPasses)
        stop("In dHMMo: probTrans was not specified correctly.  Probabilities in each row (second dimension) must sum to 1.")
      if(!obsCheckPasses)
        stop("In dHMMo: probObs was not specified correctly. Probabilities in each row must sum to 1.")
    }
    pi <- init # State probabilities at time t=1
    logL <- 0
    nObsClasses <- dim(probObs)[2]
    for (t in 1:len) {
      if (x[t] > nObsClasses | x[t] < 1) stop("In dHMMo: Invalid value of x[t].")
      Zpi <- probObs[,x[t],t] * pi # Vector of P(state) * P(observation class x[t] | state)
      sumZpi <- sum(Zpi)    # Total P(observed as class x[t])
      logL <- logL + log(sumZpi) * size  # Accumulate log probabilities through timeÃ
      if (t != len) pi <- ((Zpi %*% probTrans) / sumZpi)[1, ] # State probabilities at t+1
    }
    returnType(double())
    if (log) return(logL)
    return(exp(logL))
  }
)

rHMMopooled <- nimbleFunction(
  run = function(n = integer(),    ## Observed capture (state) history
                 init = double(1),
                 probObs = double(3),
                 probTrans = double(2),
                 len = double(0, default = 0),
                 size = double(0),
                 checkRowSums = double(0, default = 1)) {
  returnType(double(1))
  if (dim(probObs)[1] != dim(probTrans)[1]) stop("In rHMMo: Number of cols in probObs must equal number of cols in probTrans.")
  if (dim(probTrans)[1] != dim(probTrans)[2]) stop("In rHMMo: probTrans must be a square matrix.")
  if (dim(probObs)[3] != len) {
    if (dim(probObs)[3] == 1) stop("In rHMMo: Time dimension of probObs must match length of data. Did you mean rHMM?")
    stop("In rHMMo: Length of time dimension of probObs must match length of data.")
  }
  if (abs(sum(init) - 1) > 1e-6) stop("In rHMMo: Initial probabilities must sum to 1.")

  if (checkRowSums) {
    transCheckPasses <- TRUE
    for (i in 1:dim(probTrans)[1]) {
      thisCheckSum <- sum(probTrans[i,])
      if (abs(thisCheckSum - 1) > 1e-6) {
        ## Compilation doesn't support more than a simple string for stop()
        ## so we provide more detail using a print().
        print("In rHMMo: Problem with sum(probTrans[i,]) with i = ", i, ". The sum should be 1 but is ", thisCheckSum)
        transCheckPasses <- FALSE
      }
    }
    obsCheckPasses <- TRUE
    for (i in 1:dim(probObs)[1]) {
      for (k in 1:dim(probObs)[3]) {
        thisCheckSum <- sum(probObs[i,,k])
        if (abs(thisCheckSum - 1) > 1e-6) {
          print("In rHMMo: Problem with sum(probObs[i,,k]) with i = ", i, " k = ", k, ". The sum should be 1 but is ", thisCheckSum)
          obsCheckPasses <- FALSE
        }
      }
    }
    if(!(transCheckPasses | obsCheckPasses))
      stop("In rHMMo: probTrans and probObs were not specified correctly.  Probabilities in each row (second dimension) must sum to 1.")
    if(!transCheckPasses)
      stop("In rHMMo: probTrans was not specified correctly.  Probabilities in each row (second dimension) must sum to 1.")
    if(!obsCheckPasses)
      stop("In rHMMo: probObs was not specified correctly. Probabilities in each row must sum to 1.")
  }

  ans <- numeric(len)

  probInit <- init
  trueInit <- 0

  r <- runif(1, 0, 1)
  j <- 1
  while (r > sum(probInit[1:j])) j <- j + 1
  trueInit <- j

  trueState <- trueInit
  for (i in 1:len) {
    # Transition to a new true state
    r <- runif(1, 0, 1)
    j <- 1
    while (r > sum(probTrans[trueState, 1:j])) j <- j + 1
    trueState <- j

    # Detect based on the true state
    r <- runif(1, 0, 1)
    j <- 1
    while (r > sum(probObs[trueState, 1:j, i])) j <- j + 1
    ans[i] <- j

  }

  return(ans)
})

assign('dHMMopooled', dHMMopooled, .GlobalEnv)
assign('rHMMopooled', rHMMopooled, .GlobalEnv)

# code
hmm.survival <- nimbleCode({
  phi ~ dunif(0, 1) # prior survival
  p ~ dunif(0, 1) # prior detection
  # likelihood
  delta[1] <- 1          # Pr(alive t = 1) = 1
  delta[2] <- 0          # Pr(dead t = 1) = 0
  gamma[1,1] <- phi      # Pr(alive t -> alive t+1)
  gamma[1,2] <- 1 - phi  # Pr(alive t -> dead t+1)
  gamma[2,1] <- 0        # Pr(dead t -> alive t+1)
  gamma[2,2] <- 1        # Pr(dead t -> dead t+1)
  omega[1,1,1] <- 0        # Pr(alive first -> non-detected first)
  omega[1,2,1] <- 1        # Pr(alive first -> detected first)
  omega[2,1,1] <- 1        # Pr(dead first -> non-detected first)
  omega[2,2,1] <- 0        # Pr(dead first -> detected first)
  for (t in 2:5){
    omega[1,1,t] <- 1 - p    # Pr(alive t -> non-detected t)
    omega[1,2,t] <- p        # Pr(alive t -> detected t)
    omega[2,1,t] <- 1        # Pr(dead t -> non-detected t)
    omega[2,2,t] <- 0        # Pr(dead t -> detected t)
  }
  for (i in 1:N){
    y[i,1:5] ~ dHMMopooled(init = delta[1:2], # vector of initial state probabilities
                           probObs = omega[1:2,1:2,1:5], # observation matrix
                           probTrans = gamma[1:2,1:2], # transition matrix
                           len = 5, # nb of sampling occasions
                           checkRowSums = 0, # skip validity checks
                           size = size[i]) # number of individuals with encounter history i

  }
})

y_pooled <- y %>% 
  as_tibble() %>% 
  group_by_all() %>% # group
  summarise(size = n()) %>% # count
  relocate(size) %>% # put size in front
  arrange(-size) %>% # sort along size
  as.matrix()
y_pooled

my.constants <- list(N = nrow(y_pooled), T = 5, size = y_pooled[,'size'])
my.data <- list(y = y_pooled[,-1] + 1) # delete size from dataset
initial.values <- function() list(phi = runif(1,0,1),
                                  p = runif(1,0,1))
parameters.to.save <- c("phi", "p")
n.iter <- 5000
n.burnin <- 1000
n.chains <- 2

start_time <- Sys.time()
mcmc.output <- nimbleMCMC(code = hmm.survival,
                          constants = my.constants,
                          data = my.data,
                          inits = initial.values,
                          monitors = parameters.to.save,
                          niter = n.iter,
                          nburnin = n.burnin,
                          nchains = n.chains)
end_time <- Sys.time()
end_time - start_time

MCMCsummary(mcmc.output, round = 2)
```


## Decoding after marginalization 

### Theory

### Implementation

```{r}
# getViterbi() returns sequence of states that most likely generated sequence of observations
# adapted from https://github.com/vbehnam/viterbi
getViterbi <- function(Omega, Gamma, delta, y) {
# Omega: transition matrix
# Gamma: observation matrix
# delta: vector of initial state probabilities
# y: observed sequence of detections and non-detections
  
# get number of states and sampling occasions
N <- nrow(Gamma)
T <- length(y)
  
# nu is the corresponding likelihood
nu <- matrix(0, nrow = N, ncol = T)
# zz contains the most likely states up until this point
zz <- matrix(0, nrow = N, ncol = T)
firstObs <- y[1]
  
# fill in first columns of both matrices
#nu[,1] <- initial * emission[,firstObs]
#zz[,1] <- 0
nu[,1] <- c(1,0) # initial = (1, 0) * emission[,firstObs] = (1, 0)
zz[,1] <- 1 # alive at first occasion

for (i in 2:T) {
    for (j in 1:N) {
      obs <- y[i]
      # initialize to -1, then overwritten by for loop coz all possible values are >= 0
      nu[j,i] <- -1
      # loop to find max and argmax for k
      for (k in 1:N) {
        value <- nu[k,i-1] * Gamma[k,j] * Omega[j,obs]
        if (value > nu[j,i]) {
          # maximizing for k
          nu[j,i] <- value
          # argmaximizing for k
          zz[j,i] <- k
        }
      }
    }
  }
  # mlp = most likely path
  mlp <- numeric(T)
  # argmax for stateSeq[,T]
  am <- which.max(nu[,T])
  mlp[T] <- zz[am,T]
  
  # backtrace using backpointers
  for (i in T:2) {
    zm <- which.max(nu[,i])
    mlp[i-1] <- zz[zm,i]
  }
  return(mlp)
}

delta # Vector of initial states probabilities
## [1] 1 0
Gamma # Transition matrix
##      [,1] [,2]
## [1,]  0.8  0.2
## [2,]  0.0  1.0
Omega # Observation matrix
##      [,1] [,2]
## [1,]  0.4  0.6
## [2,]  1.0  0.0
getViterbi(Omega = Omega, 
           Gamma = Gamma, 
           delta = delta, 
           y = y[15,] + 1)
```

### Compute first, average after

```{r}
library(nimble)
library(tidyverse)

phi <- c(mcmc.output$chain1[,'phi'], mcmc.output$chain2[,'phi'])
p <- c(mcmc.output$chain1[,'p'], mcmc.output$chain2[,'p'])

niter <- length(p)
T <- 5
res <- matrix(NA, nrow = nrow(y), ncol = T)
for (i in 1:nrow(y)){
  res_mcmc <- matrix(NA, nrow = niter, ncol = T)
  for (j in 1:niter){
    # Initial states
    delta <- c(1, 0)
    # Transition matrix
    transition <- matrix(NA, 2, 2)
    transition[1,1] <- phi[j]      # Pr(alive t -> alive t+1)
    transition[1,2] <- 1 - phi[j]  # Pr(alive t -> dead t+1)
    transition[2,1] <- 0        # Pr(dead t -> alive t+1)
    transition[2,2] <- 1        # Pr(dead t -> dead t+1)
    # Observation matrix 
    emission <- matrix(NA, 2, 2)
    emission[1,1] <- 1 - p[j]      # Pr(alive t -> non-detected t)
    emission[1,2] <- p[j]          # Pr(alive t -> detected t)
    emission[2,1] <- 1          # Pr(dead t -> non-detected t)
    emission[2,2] <- 0          # Pr(dead t -> detected t)
    res_mcmc[j,1:T] <- getViterbi(emission, transition, delta, y[i,] + 1)
  }
  res[i, 1:length(y[1,])] <- apply(res_mcmc, 2, median)
}

df <- expand.grid(X = 1:nrow(z), Y = 1:ncol(z))
df$Z <- as_factor(c(z - res))
myPallette <- RColorBrewer::brewer.pal(name = "RdBu", n = 4)
df %>% 
  ggplot() + 
  aes(x = X, y = Y, fill = Z) + 
  geom_tile() +
  scale_fill_manual(values = c("0" = "#9597f0", 
                               "1" = "#d4b4f6", 
                               "-1" = "#ff688c"),
                    labels = c("0" = "Decoding is correct", 
                               "1" = "Dead is decoded alive", 
                               "-1" = "Alive is decoded dead"))+
  labs(x = "individuals", y = "winters", fill = NULL) +
  theme_light()
```

### Average first, compute after

```{r}
library(nimble)
library(tidyverse)

phi <- c(mcmc.output$chain1[,'phi'], mcmc.output$chain2[,'phi'])
p <- c(mcmc.output$chain1[,'p'], mcmc.output$chain2[,'p'])

# Initial states
delta <- c(1, 0)
# Transition matrix
transition <- matrix(NA, 2, 2)
transition[1,1] <- mean(phi)      # Pr(alive t -> alive t+1)
transition[1,2] <- 1 - mean(phi)  # Pr(alive t -> dead t+1)
transition[2,1] <- 0              # Pr(dead t -> alive t+1)
transition[2,2] <- 1              # Pr(dead t -> dead t+1)
# Observation matrix 
emission <- matrix(NA, 2, 2)
emission[1,1] <- 1 - mean(p)      # Pr(alive t -> non-detected t)
emission[1,2] <- mean(p)          # Pr(alive t -> detected t)
emission[2,1] <- 1                # Pr(dead t -> non-detected t)
emission[2,2] <- 0                # Pr(dead t -> detected t)
res <- matrix(NA, nrow = nrow(y), ncol = T)
for (i in 1:nrow(y)){
  res[i, 1:length(y[1,]) ] <- getViterbi(emission, transition, delta, y[i,] + 1)
}

df <- expand.grid(X = 1:nrow(z), Y = 1:ncol(z))
df$Z <- as_factor(c(z - res))
myPallette <- RColorBrewer::brewer.pal(name = "RdBu", n = 4)
df %>% 
  ggplot() + 
  aes(x = X, y = Y, fill = Z) + 
  geom_tile() +
  scale_fill_manual(values = c("0" = "#9597f0", 
                               "1" = "#d4b4f6", 
                               "-1" = "#ff688c"),
                    labels = c("0" = "Decoding is correct", 
                               "1" = "Dead is decoded alive", 
                               "-1" = "Alive is decoded dead"))+
  labs(x = "individuals", y = "winters", fill = NULL) +
  theme_light()
```

## Summary

## Suggested reading

