---
title: "Chapter 2: NIMBLE tutorial"
date: "last updated: `r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE, echo=FALSE, message=FALSE, warning=FALSE}
# R options
options(width = 60)
# chunk options
knitr::opts_chunk$set(
  comment = "##",
  collapse = TRUE,
  warning = FALSE,
  message = FALSE
  )
```

## Introduction

## What is NIMBLE?

## Getting started

```{r}
library(nimble)
library(tidyverse)
library(MCMCvis)

model <- nimbleCode({
  # likelihood
  survived ~ dbinom(theta, released)
  # prior
  theta ~ dunif(0, 1)
  # derived quantity
  lifespan <- -1/log(theta)
})

my.data <- list(released = 57, survived = 19)

parameters.to.save <- c("theta", "lifespan")

# init1 <- list(theta = 0.1)
# init2 <- list(theta = 0.5)
# init3 <- list(theta = 0.9)
# initial.values <- list(init1, init2, init3)

initial.values <- function() list(theta = runif(1,0,1))

n.iter <- 5000
n.burnin <- 1000
n.chains <- 3

mcmc.output <- nimbleMCMC(code = model,
                          data = my.data,
                          inits = initial.values,
                          monitors = parameters.to.save,
                          niter = n.iter,
                          nburnin = n.burnin,
                          nchains = n.chains)

str(mcmc.output)

dim(mcmc.output$chain1)

head(mcmc.output$chain1)

mean(mcmc.output$chain1[,'theta'])

quantile(mcmc.output$chain1[,'theta'], probs = c(2.5, 97.5)/100)

mcmc.output %>%
  as_tibble() %>%
  ggplot() +
  geom_histogram(aes(x = chain1[,"theta"]), color = "white") +
  labs(x = "survival probability")

MCMCsummary(object = mcmc.output, round = 2)

MCMCplot(object = mcmc.output, 
         params = 'theta')

MCMCtrace(object = mcmc.output,
          pdf = FALSE, # no export to PDF
          ind = TRUE, # separate density lines per chain
          params = "theta")

MCMCtrace(object = mcmc.output,
          pdf = FALSE,
          ind = TRUE,
          Rhat = TRUE, # add Rhat
          n.eff = TRUE, # add eff sample size
          params = "theta")

theta_samples <- c(mcmc.output$chain1[,'theta'], 
                   mcmc.output$chain2[,'theta'],
                   mcmc.output$chain3[,'theta'])

lifespan <- -1/log(theta_samples)

mean(lifespan)

quantile(lifespan, probs = c(2.5, 97.5)/100)

lifespan %>%
  as_tibble() %>%
  ggplot() +
  geom_histogram(aes(x = value), color = "white") +
  labs(x = "lifespan")
```

## Programming

### NIMBLE functions


```{r}
library(nimble)
library(MCMCvis)

computeLifespan <- nimbleFunction(
    run = function(theta = double(0)) { # type declarations
        ans <- -1/log(theta)
        return(ans)
        returnType(double(0))  # return type declaration
    } )

computeLifespan(0.8)

CcomputeLifespan <- compileNimble(computeLifespan)
CcomputeLifespan(0.8)

model <- nimbleCode({
  # likelihood
  survived ~ dbinom(theta, released)
  # prior
  theta ~ dunif(0, 1)
  # derived quantity
  lifespan <- computeLifespan(theta)
})

my.data <- list(survived = 19, released = 57)
parameters.to.save <- c("theta", "lifespan")
initial.values <- function() list(theta = runif(1,0,1))
n.iter <- 5000
n.burnin <- 1000
n.chains <- 3
mcmc.output <- nimbleMCMC(code = model,
                          data = my.data,
                          inits = initial.values,
                          monitors = parameters.to.save,
                          niter = n.iter,
                          nburnin = n.burnin,
                          nchains = n.chains)

MCMCsummary(object = mcmc.output, round = 2)
```


### Calling R/C++ functions


```{r}
library(nimble)
library(MCMCvis)

myfunction <- function(x) {
  -1/log(x)
}

Rmyfunction <- nimbleRcall(prototype = function(x = double(0)){}, 
                           Rfun = 'myfunction',
                           returnType = double(0))

model <- nimbleCode({
  # likelihood
  survived ~ dbinom(theta, released)
  # prior
  theta ~ dunif(0, 1)
  lifespan <- Rmyfunction(theta)
})

my.data <- list(survived = 19, released = 57)
parameters.to.save <- c("theta", "lifespan")
initial.values <- function() list(theta = runif(1,0,1))
n.iter <- 5000
n.burnin <- 1000
n.chains <- 3
mcmc.output <- nimbleMCMC(code = model,
                          data = my.data,
                          inits = initial.values,
                          monitors = parameters.to.save,
                          niter = n.iter,
                          nburnin = n.burnin,
                          nchains = n.chains)

MCMCsummary(object = mcmc.output, round = 2)
```


### User-defined distributions

```{r}
library(nimble)
library(MCMCvis)

# density
dmybinom <- nimbleFunction(
  run = function(x = double(0), 
                 size = double(0), 
                 prob = double(0), 
                 log = integer(0, default = 1)) {
    returnType(double(0))
    # compute binomial coefficient 
    lchoose <- lfactorial(size) - lfactorial(x) - lfactorial(size - x)
    # binomial density function
    logProb <- lchoose + x * log(prob) + (size - x) * log(1 - prob)
    if(log) return(logProb)
    else return(exp(logProb)) 
  })
# simulation using the coin flip method (p. 524 in Devroye 1986)
rmybinom <- nimbleFunction(
  run = function(n = integer(0, default = 1),
                 size = double(0),
                 prob = double(0)) {
    returnType(double(0))
    x <- 0
    y <- runif(n = size, min = 0, max = 1)
    for (j in 1:size){
      if (y[j] < prob){
        x <- x + 1
      }else{
        x <- x
      }
    }
    return(x)    
  })

assign('dmybinom', dmybinom, .GlobalEnv)
assign('rmybinom', rmybinom, .GlobalEnv)

rmybinom(n = 1, size = 5, prob = 0.1)

model <- nimbleCode({
 # likelihood
 survived ~ dmybinom(prob = theta, size = released)
 # prior
 theta ~ dunif(0, 1)
})
my.data <- list(released = 57, survived = 19)
initial.values <- function() list(theta = runif(1,0,1))
n.iter <- 5000
n.burnin <- 1000
n.chains <- 3
mcmc.output <- nimbleMCMC(code = model,
 data = my.data,
 inits = initial.values,
 niter = n.iter,
 nburnin = n.burnin,
 nchains = n.chains)

MCMCsummary(mcmc.output)
```

## Under the hood 

```{r}
library(nimble)
library(MCMCvis)

model <- nimbleCode({
  # likelihood
  survived ~ dbinom(theta, released)
  # prior
  theta ~ dunif(0, 1)
  # derived quantity
  lifespan <- -1/log(theta)
})
my.data <- list(survived = 19, released = 57)
initial.values <- list(theta = 0.5)

survival <- nimbleModel(code = model,
                        data = my.data,
                        inits = initial.values)

survival$getNodeNames()

survival$theta

survival$survived

survival$lifespan 

survival$calculate()

survival$plotGraph()

Csurvival <- compileNimble(survival)

Csurvival$theta

f <- function(par) {
    Csurvival[['theta']] <- par # assign par to theta 
    ll <- Csurvival$calculate() # update log-likelihood with par value
    return(-ll) # return negative log-likelihood
}

f(0.5)

f(0.9)

out <- optimize(f, interval = c(0,1))
round(out$minimum, 2)

survivalConf <- configureMCMC(survival)

survivalConf$addMonitors(c("lifespan"))

survivalMCMC <- buildMCMC(survivalConf)
CsurvivalMCMC <- compileNimble(survivalMCMC, project = survival)

n.iter <- 5000
n.burnin <- 1000
samples <- runMCMC(mcmc = CsurvivalMCMC, 
                   niter = n.iter,
                   nburnin = n.burnin)

head(samples)

samplesSummary(samples)
```

## MCMC samplers

### Default samplers

```{r}
library(nimble)
library(MCMCvis)

model <- nimbleCode({
  # likelihood
  survived ~ dbinom(theta, released)
  # prior
  theta ~ dunif(0, 1)
  # derived quantity
  lifespan <- -1/log(theta)
})

my.data <- list(survived = 19, released = 57)

initial.values <- list(theta = 0.5)

survival <- nimbleModel(code = model,
                        data = my.data,
                        inits = initial.values)

Csurvival <- compileNimble(survival)

survivalConf <- configureMCMC(survival)

survivalConf$addMonitors(c("lifespan"))

survivalMCMC <- buildMCMC(survivalConf)
CsurvivalMCMC <- compileNimble(survivalMCMC, project = survival)

n.iter <- 5000
n.burnin <- 1000
samples <- runMCMC(mcmc = CsurvivalMCMC, 
                   niter = n.iter,
                   nburnin = n.burnin)

survivalConf$printSamplers()

survivalConf$removeSamplers(c('theta'))
survivalConf$printSamplers()

survivalConf$addSampler(target = c('theta'),
                        type = 'slice')
survivalConf$printSamplers()

survivalMCMC2 <- buildMCMC(survivalConf)
CsurvivalMCMC2 <- compileNimble(survivalMCMC2, 
                                project = survival,
                                resetFunctions = TRUE)

samples2 <- runMCMC(mcmc = CsurvivalMCMC2, 
                    niter = n.iter,
                    nburnin = n.burnin)

samplesSummary(samples2)
```

### User-defined samplers 

```{r}
library(nimble)

my_metropolis <- nimbleFunction(
  name = 'my_metropolis', # fancy name for our MCMC sampler
  contains = sampler_BASE,
  setup = function(model, mvSaved, target, control) {
    # i) get dependencies for 'target' in 'model'
    calcNodes <- model$getDependencies(target) 
    # ii) get sd of proposal distribution
    scale <- control$scale 
  },
  run = function() {
    # (1) log-lik at current value
    initialLP <- model$getLogProb(calcNodes) 
    # (2) current parameter value
    current <- model[[target]] 
    # (3) logit transform
    lcurrent <- log(current / (1 - current))
    # (4) propose candidate value
    lproposal <- lcurrent  + rnorm(1, mean = 0, scale) 
    # (5) back-transform
    proposal <- plogis(lproposal)
    # (6) plug candidate value in model 
    model[[target]] <<- proposal 
    # (7) log-lik at candidate value
    proposalLP <- model$calculate(calcNodes)
    # (8) compute lik ratio on log scale
    lMHR <- proposalLP - initialLP 
    # (9) spin continuous spinner and compare to ratio
    if(runif(1,0,1) < exp(lMHR)) { 
      # (10) if candidate value is accepted, update current value
      copy(from = model, to = mvSaved, nodes = calcNodes, logProb = TRUE, row = 1)
    } else {
      ## (11) if candidate value is accepted, keep current value
      copy(from = mvSaved, to = model, nodes = calcNodes, logProb = TRUE, row = 1)
    }
  },
  methods = list(
    reset = function() {}
  )
)

model <- nimbleCode({
  # likelihood
  survived ~ dbinom(theta, released)
  # prior
  theta ~ dunif(0, 1)
})
my.data <- list(survived = 19, released = 57)
initial.values <- function() list(theta = runif(1,0,1))
survival <- nimbleModel(code = model, 
                        data = my.data, 
                        inits = initial.values())
Csurvival <- compileNimble(survival)
survivalConf <- configureMCMC(survival)

survivalConf$printSamplers()

survivalConf$removeSamplers(c('theta'))
survivalConf$addSampler(target = 'theta', 
                        type = 'my_metropolis', 
                        control = list(scale = 0.1)) 

survivalConf$printSamplers()

survivalMCMC <- buildMCMC(survivalConf)
CsurvivalMCMC <- compileNimble(survivalMCMC, 
                               project = survival)
samples <- runMCMC(mcmc = CsurvivalMCMC, 
                   niter = 5000, 
                   nburnin = 1000)

samplesSummary(samples)
```


```{r}
library(nimble)
library(tidyverse)
library(patchwork)

scale <- 0.1
Rmodel <- nimbleModel(code = model, data = my.data, inits = initial.values())
conf <- configureMCMC(Rmodel, monitors = c('theta'), print = FALSE)
conf$removeSamplers(c('theta'))
conf$addSampler(target = 'theta', type = 'my_metropolis', control = list(scale = scale))
Rmcmc <- buildMCMC(conf)
out <- compileNimble(list(model = Rmodel, mcmc = Rmcmc))
Cmcmc <- out$mcmc
samples_sd01 <- runMCMC(Cmcmc, niter = 10000, nburnin = 9000, progressBar = FALSE)
scale <- 1
Rmodel <- nimbleModel(code = model, data = my.data, inits = initial.values())
conf <- configureMCMC(Rmodel, monitors = c('theta'), print = FALSE)
conf$removeSamplers(c('theta'))
conf$addSampler(target = 'theta', type = 'my_metropolis', control = list(scale = scale))
Rmcmc <- buildMCMC(conf)
out <- compileNimble(list(model = Rmodel, mcmc = Rmcmc))
Cmcmc <- out$mcmc
samples_sd1 <- runMCMC(Cmcmc, niter = 10000, nburnin = 9000, progressBar = FALSE)
scale <- 10
Rmodel <- nimbleModel(code = model, data = my.data, inits = initial.values())
conf <- configureMCMC(Rmodel, monitors = c('theta'), print = FALSE)
conf$removeSamplers(c('theta'))
conf$addSampler(target = 'theta', type = 'my_metropolis', control = list(scale = scale))
Rmcmc <- buildMCMC(conf)
out <- compileNimble(list(model = Rmodel, mcmc = Rmcmc))
Cmcmc <- out$mcmc
samples_sd10 <- runMCMC(Cmcmc, niter = 10000, nburnin = 9000, progressBar = FALSE)

plot01 <- samples_sd01 %>%
  as_tibble() %>%
  ggplot() + 
  aes(x = 9001:10000, y = theta) +
  geom_line() + 
  labs(x = "iterations", title = "scale = 0.1")
plot1 <- samples_sd1 %>%
  as_tibble() %>%
  ggplot() + 
  aes(x = 9001:10000, y = theta) +
  geom_line() + 
  labs(x = "iterations", title = "scale = 1")
plot10 <- samples_sd10 %>%
  as_tibble() %>%
  ggplot() + 
  aes(x = 9001:10000, y = theta) +
  geom_line() + 
  labs(x = "iterations", title = "scale = 10")

plot01 + plot1 + plot10
```


## Tips and tricks 

### Precision vs standard deviation 

### Indexing 

### Faster compilation

```{r}
library(nimble)

model <- nimbleCode({
  survived ~ dbinom(theta, released)
  theta ~ dunif(0, 1)
})
my.data <- list(survived = 19, released = 57)
initial.values <- function() list(theta = runif(1,0,1))
survival <- nimbleModel(code = model, 
                        data = my.data, 
                        inits = initial.values(),
                        calculate = FALSE) # first tip
Csurvival <- compileNimble(survival)
survivalConf <- configureMCMC(survival)

survivalMCMC <- buildMCMC(survivalConf, useConjugacy = FALSE) # second tip
CsurvivalMCMC <- compileNimble(survivalMCMC, 
                               project = survival)
samples <- runMCMC(mcmc = CsurvivalMCMC, 
                   niter = 5000, 
                   nburnin = 1000)

samplesSummary(samples)
```

### Updating MCMC chains 

```{r}
library(nimble)

model <- nimbleCode({
  survived ~ dbinom(theta, released)
  theta ~ dunif(0, 1)
})
my.data <- list(survived = 19, released = 57)
initial.values <- function() list(theta = runif(1,0,1))
survival <- nimbleModel(code = model, 
                        data = my.data, 
                        inits = initial.values(),
                        calculate = FALSE) # first tip
Csurvival <- compileNimble(survival)
survivalConf <- configureMCMC(survival)

survivalMCMC <- buildMCMC(survivalConf, useConjugacy = FALSE) # second tip
CsurvivalMCMC <- compileNimble(survivalMCMC, 
                               project = survival)
samples <- runMCMC(mcmc = CsurvivalMCMC, 
                   niter = 5000, 
                   nburnin = 1000)

samplesSummary(samples)

niter_ad <- 6000
CsurvivalMCMC$run(niter_ad, reset = FALSE)

more_samples <- as.matrix(CsurvivalMCMC$mvSamples)
samplesSummary(more_samples)
```


### Updating MCMC chains

```{r}
library(nimble)

model <- nimbleCode({
  survived ~ dbinom(theta, released)
  theta ~ dunif(0, 1)
})

my.data <- list(survived = 19, released = 57)

initial.values <- function() list(theta = runif(1,0,1))

survival <- nimbleModel(code = model, 
                        data = my.data, 
                        inits = initial.values(),
                        calculate = FALSE) # first tip

# first call to nimbleMCMC()
mcmc.output1 <- nimbleMCMC(code = model,
                           data = my.data,
                           inits = initial.values,
                           niter = 5000,
                           nburnin = 1000,
                           nchains = 3,
                           summary = TRUE,
                           setSeed = 123)

# second call to nimbleMCMC()
mcmc.output2 <- nimbleMCMC(code = model,
                           data = my.data,
                           inits = initial.values,
                           niter = 5000,
                           nburnin = 1000,
                           nchains = 3,
                           summary = TRUE,
                           setSeed = 123)

mcmc.output1$summary$all.chains

mcmc.output2$summary$all.chains
```

### Parallelization 
 
```{r}
library(nimble)
library(MCMCvis)
library(parallel)

nbcores <- detectCores() - 1
my_cluster <- makeCluster(nbcores)

workflow <- function(seed, data) {
  
  library(nimble)
  
  model <- nimbleCode({
    # likelihood
    survived ~ dbinom(theta, released)
    # prior
    theta ~ dunif(0, 1)
  })
  
  set.seed(123) # for reproducibility
  initial.values <- function() list(theta = runif(1,0,1))
  
  survival <- nimbleModel(code = model, 
                          data = data, 
                          inits = initial.values())
  Csurvival <- compileNimble(survival)
  survivalMCMC <- buildMCMC(Csurvival)
  CsurvivalMCMC <- compileNimble(survivalMCMC)
  
  samples <- runMCMC(mcmc = CsurvivalMCMC, 
                     niter = 5000, 
                     nburnin = 1000,
                     setSeed = seed)
  
  return(samples)
}

output <- parLapply(cl = my_cluster, 
                    X = c(2022, 666),
                    fun = workflow, 
                    data = list(survived = 19, released = 57))

stopCluster(my_cluster)

str(output)

MCMCsummary(output)
```


### Incomplete initialization 

```{r}
library(nimble)

model <- nimbleCode({
  survived ~ dbinom(theta, released)
  theta ~ dunif(0, 1)
})

survival <- nimbleModel(code = model, 
                        data = list(survived = 19, released = 57))

survival$initializeInfo()

survival$theta <- 0.5 # assign initial value to theta
survival$calculate() 

Csurvival <- compileNimble(survival)
survivalMCMC <- buildMCMC(Csurvival)

CsurvivalMCMC <- compileNimble(survivalMCMC)

samples <- runMCMC(mcmc = CsurvivalMCMC, 
                   niter = 5000, 
                   nburnin = 1000)

samplesSummary(samples)
```

### Vectorization

## Summary

## Suggested reading

