---
title: "wolf and ih"
date: "last updated: `r Sys.Date()`"
output: html_document
---


```{r setup, include=FALSE, echo=FALSE}
options(htmltools.dir.version = FALSE)
knitr::opts_chunk$set(comment = "", message = TRUE, warning = TRUE)
```

## Introduction

How to incorporate individual heterogeneity in capture-recapture models using finite mixtures. 

## Pre-requisites

```{r}
library(tidyverse)
theme_set(theme_light(base_size = 14))
library(nimble)
library(MCMCvis)
```

## Data

```{r}
y <- as.matrix(read.table("wolf.txt"))
# 1 seen
# 0 not seen
dim(y)
```

## Model

Let's write the model. Only thing to notice is that we have two alive states to represent two classes of individuals. We choose to have the detection probabilities dependent on the states and constant survival. 
```{r}
hmm.phipmix <- nimbleCode({
  
  # priors
  phi ~ dunif(0, 1) # prior survival
  p1 ~ dunif(0, 1) # prior detection
  p2 ~ dunif(0, 1) # prior detection
  pi ~ dunif(0, 1) # prob init state 1

  # HMM ingredients
  gamma[1,1] <- phi      # Pr(alive 1 t -> alive 1 t+1)
  gamma[1,2] <- 0        # Pr(alive 1 t -> alive 2 t+1) // no transition
  gamma[1,3] <- 1 - phi  # Pr(alive t -> dead t+1)
  gamma[2,1] <- 0        # Pr(alive 1 t -> alive 1 t+1) // no transition
  gamma[2,2] <- phi      # Pr(alive 1 t -> alive 2 t+1) 
  gamma[2,3] <- 1 - phi  # Pr(alive t -> dead t+1)
  gamma[3,1] <- 0        # Pr(dead t -> alive 1 t+1)
  gamma[3,2] <- 0        # Pr(dead t -> alive 1 t+1)
  gamma[3,3] <- 1        # Pr(dead t -> dead t+1)
  delta[1] <- pi         # Pr(alive t = 1) = pi
  delta[2] <- 1 - pi     # Pr(alive t = 1) = 1 - pi
  delta[3] <- 0          # Pr(dead t = 1) = 0
  omega[1,1] <- 1 - p1   # Pr(alive state 1 t -> non-detected t)
  omega[1,2] <- p1       # Pr(alive state 1 t -> detected t)
  omega[2,1] <- 1 - p2   # Pr(alive state 2 t -> non-detected t)
  omega[2,2] <- p2       # Pr(alive state 2 t -> detected t)
  omega[3,1] <- 1        # Pr(dead t -> non-detected t)
  omega[3,2] <- 0        # Pr(dead t -> detected t)
  omega.init[1,1] <- 0   # Pr(alive state 1 t -> non-detected t)
  omega.init[1,2] <- 1   # Pr(alive state 1 t -> detected t)
  omega.init[2,1] <- 0   # Pr(alive state 2 t -> non-detected t)
  omega.init[2,2] <- 1   # Pr(alive state 2 t -> detected t)
  omega.init[3,1] <- 1   # Pr(dead t -> non-detected t)
  omega.init[3,2] <- 0   # Pr(dead t -> detected t)
  
  # likelihood
  for (i in 1:N){
    z[i,first[i]] ~ dcat(delta[1:3])
    y[i,first[i]] ~ dcat(omega.init[z[i,first[i]], 1:2])
    for (j in (first[i]+1):K){
      z[i,j] ~ dcat(gamma[z[i,j-1], 1:3])
      y[i,j] ~ dcat(omega[z[i,j], 1:2])
    }
  }
})
```

## Ingredients

Get the date of first capture. 
```{r}
first <- apply(y, 1, function(x) min(which(x != 0)))
```

Constants in a list. 
```{r}
my.constants <- list(N = nrow(y), 
                     K = ncol(y), 
                     first = first)
```

Data in a list. 
```{r}
my.data <- list(y = y + 1)
```

Initial values. 
```{r}
zinit <- y
for (i in 1:nrow(y)) {
  for (j in first[i]:ncol(y)) {
    if (j == first[i]) zinit[i,j] <- which(rmultinom(1, 1, c(1/2,1/2))==1) # pick alive state
    if (j > first[i]) zinit[i,j] <- zinit[i,j-1] # because no transition, state remains the same
  }
}
zinit <- as.matrix(zinit)
initial.values <- function() list(phi = runif(1,0,1),
                                  p1 = runif(1,0,1),
                                  p2 = runif(1,0,1),
                                  pi = runif(1,0,1),
                                  z = zinit)
```

Parameters to be monitored. 
```{r}
parameters.to.save <- c("phi", "p1", "p2", "pi")
```

MCMC details. 
```{r}
n.iter <- 10000
n.burnin <- 5000
n.chains <- 2
```

## Fitting

Run NIMBLE:
```{r}
mcmc.phipmix <- nimbleMCMC(code = hmm.phipmix, 
                           constants = my.constants,
                           data = my.data,              
                           inits = initial.values,
                           monitors = parameters.to.save,
                           niter = n.iter,
                           nburnin = n.burnin, 
                           nchains = n.chains)
```

## Post-processing

Numerical summaries. 
```{r}
MCMCsummary(mcmc.phipmix, round = 2)
```

Let's have a look to the trace and posterior distribution of the proportion of individuals in each class, as well as the detection and survival probabilities. 
```{r}
MCMCtrace(mcmc.phipmix, pdf = FALSE, params = c("pi", "p1", "p2", "phi"))
```


